Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-10-14 Dail_Mad_ext.tex
rough simulation results plus

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect
detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential, Ricker, etc $\dots$), zero-inflation, and
random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} or \textbf{JAGS} is also provided.
A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
}

@@

\textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
Gompertz model, immigration, open population point count models,
random observer effects, range, Ricker model, zero-inflated

@@

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\grave{o}_t$.
The growth model may be density-dependent, as in the case of the
Ricker model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcom of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are among the most widespread approaches for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the use of Gaussian distribution for the random
effects, (2) the model for observation error has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
paramters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation (i.e., $[Y|\Omega]$ is interpretted
as the probability distribution of the random variable $Y$ given the
parameter $\Omega$). The three probability distributions required for a
state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
%  so these equations are confusing to me.  How widely used is this
%  notation in ecology?  Will others likely understand this?  Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
And how could $\bm{\theta}$ be
estimated given that there is only a single observation available?
Often, these problems are simply assumed away. For
example, sometimes researchers assume that there is no process variation in
the first year, although this seems unjustified. Alternatively, the
population may be assumed to be at equilibrium such that the expected
value of $N_t$ is constant through time. Although this
makes the parameters $\bm{\theta}$ identifiable, assuming equlibrium
defeats the objective of many studies of population dynamics, namely
determining why a population varies over time.

@@

In this paper, we focus on the model of
\citet[henceforth the DM model;]{dail_madsen:2011}
that simultaneous resolves each of the abovementioned problems with
traditional state-space models, and is designed for simple count
data.
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficiencies with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accomodate classical models of population growth and to
handle several features common to ecological
time-series. Specifically, we describe methods for accomodating excess
zeros and nuisance variables such as random observer effects. Both
frequentist and Bayesian methods of inference are discussed, and code
for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accomodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of  the paper is to provide ecologists with flexible and accessible means of
adressing important questions related to the variation of abundance in
space and time.
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.

\begin{comment}
  The DM model allows for inference about spatial and temporal
  dynamics in abundance while accounting for observation error using
  only spatially and temporally replicated count data.  Their model is
  an extension of the closed-population $N$-mixture model
  \citep{royle:2004biom}, which was designed specifically to address
  the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
  assumption of population closure and includes explicit parameters
  describing population change as a first-order Markovian
  process. [more details]

  [applications / potential etc... ] The DM model is relevant to a
  vast number of problems confronting basic and applied ecological
  research. Answering questions about climate change etc... require
  spatially and temporally extensive datasets such as result from the
  North American Breeding Bird Survey (BBS)
  \citep{robbins_etal:1986}. However, most of the datasets of the
  appropriate scale are collected by volunteers and using protocols
  that are not amenable to traditional approaches of modeling
  abundance and detection probability based on standard
  capture-recapture methods. Instead they have, at best, simple count
  data for which, when it can be interpreted as counts of unique
  individuals, the DM model can be applied.. Yadi ya

@@

  In this paper, we demonstrate how the DM model resolves each of
  these four limitations of conventional state-space models, and we
  extend the DM model in several important ways. First, we demonstrate
  how classical population models can be embedded within the DM
  model. Then we present methods for handling excess zeros as well as
  nuisance parameters such as random observer effects on detection
  probability. A simulation study
\end{comment}

@@

The DM model requires count data collected at $R$ sites, each of which
is surveyed on $T$ primary sampling periods. A site is ideally a
well-defined region of a study area, such as a wetland or a patch of
early-successional habitat, although it could be an arbitrarily
defined region such as a randomly located
survey plots. The timeframe of the study is arbitrary, but in general
it will be sufficiently long to allow for temporal variation in
abundance. For instance, a primary sampling period could be a one
month long breeding season, which could be surveyed once per year for
$T$ years. If the a site is surveyed on $J$ occassions during a
primary period, these are called secondary sampling occasions.
In the case where no secondary sampling was conducted, i.e. $J=1$,
let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
$i$ and primary period $t$. If secondary sampling periods were used,
an extra dimension is added so tha twe have $X_{ijt}:
j=1,\hdots,J$. Typically, we $X_{it}$ will be lower than the actual
quantitiy of interest, abundance,
denoted $X_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{it} \equiv N_{it}$.

@@

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during year
1.
The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of overdispersion
determined by the parameter $\alpha$.
In the Poisson case, the eqivalent expression of
$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
\begin{equation}
  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
  \label{eq:N1br}
\end{equation}

Regardless of the specified distribution, the model for initial
abundance has two distinguising features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one might consider the influence of some environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 + \beta^{\Lambda}_1
x_{i}$. The second important point is that the spatial replicates resolve the
problem of parameter non-identifiability that are common to standard
state-space models because, as demonstrated by \citet{royle:2004biom},
process variation and observation error can be estimated from
spatially-replicated count data. Hence, the first component of the
model addresses both the issues of spatial inference and parameter
identifiability discussed previously.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although
higher order processes are also possible.  \citet{dail_madsen:2011}
considered several models to describe the temporal dynamics; however,
in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering the
population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$ is
the recruitment rate (which can depend on $N_{it-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
autoregressive model, $G_{it} \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accomodated, for example using
logit- and log-linear models respectively.

To make the connection between conventional state-space models and the
DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
Eq.~\ref{eq:ss2b}, with an expression derived from
Eq.~\ref{eq:Nt}. This requires summing over all possible values of $S$
and $G$, which is accomplished using the discrete convolution:
\begin{equation}
[N_{it}|N_{it-1}, \omega, \gamma] =
  \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
  \binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}} \times
  \frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}
  \label{eq:P}
\end{equation}
where here we assumed no dependenced of $\gamma$ on $N_{it-1}$.

In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate (under
the geometric growth model). This ability to directly estimate
demographic parameters from simple count data is one of the DM model's
most appearling features. However,
in the more common scenario when immigration and emigration occur,
$\omega$ and $\gamma$ no longer can be intepretted as vital
rates. Rather, $\omega$ is the probability of surviving and not
emigrating, and
$\gamma$ is the sum of the birth rate and the immigration rate,
i.e. the recruitment rate. In such cases, we propose
replacing Eq.~\ref{eq:Nt} with standard
population models.

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model, i.e. the
model for abundance dynamics. This state
model has allowed us to replace the initial abundance distribution and
transition models of conventional state-space models, with
alternatives that (1) allow for modeling spatial variation in
abundance, (2) respect the discrete nature of the counts, and (3) have
parameters that are estimable. The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be more
easily interpretted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection, and the one
considered in the original formulation, is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
or equivalently,
\begin{equation}
  [X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}} \bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
  \label{eq:1}
\end{equation}
Under this model, $\bm{p}$ are interpretted as detection probability
parameters...

%%%% Save for discussion?
%Surprsingly detection probability, $p$, can be estimated without
%secondary sampling occasions. However, this is only possible
%if the parameteric assumptions of the population dynamics are met
%exactly. In practice, it is preferable to obtain direct information
%about $p$ using secondary sampling occassions or some other auxiliary
%data, \hl{as is demonstrated in section....}

@@

Maximum likelihood estimation of parameters in models with discrete random
effects requires removing them by
summation. For Markovian models, this can be done recursively, making
the likelihood tractible. A general expression of the likelihood can be written:
\begin{multline}
  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
%  \prod_{t=T}^2 \left\{
  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
  \right\} \right\}
  \label{eq:like}
\end{multline}
where $\infty$ is replaced by a finite upper bound $K$, which should
be high enough such that the MLEs are not affected by increasint it
further. This likelihood can be maximized numerically using the \textbf{R}
package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples are
given in the Appendix.

@@

We thank J. Andrew Royle and T. Scott Sillett for helpful comments
and suggestions.

>>>>

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential, Ricker-logistic, etc $\dots$), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} or \textbf{JAGS} is also provided.
A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
}

@@

\textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
Gompertz-logistic model, immigration, open population point count models,
random observer effects, range, Ricker-logistic model, zero-inflated

@@

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\grave{o}_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcome of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are among the most widespread approaches for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the use of Gaussian distribution for the random
effects, (2) the model for observation error has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
parameters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation (i.e., $[Y|\Omega]$ is interpreted
as the probability distribution of the random variable $Y$ given the
parameter $\Omega$). The three probability distributions required for a
state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
\label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
And how could $\bm{\theta}$ be estimated given that there is only a single observation
available? Often, these problems are simply assumed away. For
example, sometimes researchers assume that there is no process variation in
the first year, although this seems unjustified. Alternatively, the
population may be assumed to be at equilibrium such that the expected
value of $N_t$ is constant through time. Although this 
makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
defeats the objective of many studies of population dynamics, namely
determining why a population varies over time.

@@

In this paper, we focus on the model of \citet[henceforth the DM model;]{dail_madsen:2011}
that simultaneous resolves each of the aforementioned problems with
traditional state-space models, and is designed for simple countdata.
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficiencies with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time-series. Specifically, we describe methods for accommodating
excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of the paper is to provide ecologists with flexible and accessible means of
addressing important questions related to the variation of abundance in
space and time.
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.
\begin{comment}
  The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error
using only spatially and temporally replicated count data. Their model
is an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to
address the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
assumption of population closure and includes explicit
parameters describing population change as a first-order Markovian
  process. [more details]

[applications / potential etc... ] The DM model is relevant to a
vast number of problems confronting basic and applied ecologicalresearch. Answering questions about climate change etc...
require
spatially and temporally extensive datasets such as result from
the
  North American Breeding Bird Survey (BBS)
\citep{robbins_etal:1986}. However, most of the datasets of theappropriate scale are collected by volunteers and using
protocols
  that are not amenable to traditional approaches of modeling
  abundance and detection probability based on standard
capture-recapture methods. Instead they have, at best, simple
count
data for which, when it can be interpreted as counts of unique  individuals, the DM model can be applied.. Yadi ya

@@

In this paper, we demonstrate how the DM model resolves each ofthese four limitations of conventional state-space models, and
we
extend the DM model in several important ways. First, we
demonstrate
  how classical population models can be embedded within the DM
model. Then we present methods for handling excess zeros as well
as
nuisance parameters such as random observer effects on detection  probability. A simulation study
\end{comment}

@@

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. A site is ideally a
well-defined region of a study area, such as a wetland or a
patch of early-successional habitat, although it could be an arbitrarily
defined region such as a randomly located survey plots. The time
frame of the study is arbitrary, but in general
it will be sufficiently long to allow for temporal variation in
abundance. For instance, a primary sampling period could be a
one month long breeding season, which could be surveyed once per
year for $T$ years. If the a site is surveyed on $J$ occasions during a
primary period, these are called secondary sampling occasions.
In the case where no secondary sampling was conducted, i.e.
$J=1$, let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. If secondary sampling periods were
used, an extra dimension is added so that we have $X_{ijt}:
j=1,\hdots,J$. Typically, we $X_{it}$ will be lower than the
actual quantity of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{it} \equiv N_{it}$.

@@

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
In the Poisson case, the equivalent expression of
$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
\begin{equation}
[N_{i1} | \Lambda] =
\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
  \label{eq:N1br}
\end{equation}

Regardless of the specified distribution, the model for initial
abundance has two distinguishing features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one might consider the influence of some environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 +
\beta^{\Lambda}_1
x_{i}$. The second important point is that the spatial
replicates resolve the
problem of parameter non-identifiability that are common to
standard state-space models because, as demonstrated by
\citet{royle:2004biom},
process variation and observation error can be estimated from
spatially-replicated count data. Hence, the first component of
the model addresses both the issues of spatial inference and
parameter identifiability discussed previously.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{it-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{it-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{it} \sim
\mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

To make the connection between conventional state-space models
and the DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
Eq.~\ref{eq:ss2b}, with an expression derived from
Eq.~\ref{eq:Nt}. This requires summing over all possible values
of $S$ and $G$, which is accomplished using the discrete convolution:
\begin{equation}
[N_{it}|N_{it-1}, \omega, \gamma] =
  \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
\binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}}
\times
\frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}  \label{eq:P}
\end{equation}
where here we assumed no dependence of $\gamma$ on $N_{it-1}$.

In the absence of movement, $\omega$ is exactly the probability of surviving
from year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate
(under the geometric growth model). This ability to directly estimate
demographic parameters from simple count data is one of the DM
model's most appearling features. However,
in the more common scenario when immigration and emigration
occur, $\omega$ and $\gamma$ no longer can be interpreted as vital
rates. Rather, $\omega$ is the probability of surviving and not
emigrating, and $\gamma$ is the sum of the birth rate and the immigration rate,
i.e. the recruitment rate. In such cases, we propose
replacing Eq.~\ref{eq:Nt} with standard population models.

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
i.e. the model for abundance dynamics. This state
model has allowed us to replace the initial abundance
distribution and transition models of conventional state-space models, with
alternatives that (1) allow for modeling spatial variation in
abundance, (2) respect the discrete nature of the counts, and
(3) have parameters that are estimable. The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection, and the
one considered in the original formulation, is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
or equivalently,
\begin{equation}
[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
  \label{eq:1}
\end{equation}
Under this model, $\bm{p}$ are interpretted as detection
probability
parameters...

%%%% Save for discussion?
%Surprisingly detection probability, $p$, can be estimated without
%secondary sampling occasions. However, this is only possible
%if the parametric assumptions of the population dynamics are met
%exactly. In practice, it is preferable to obtain direct information
%about $p$ using secondary sampling occasions or some other auxiliary
%data, \hl{as is demonstrated in section....}

@@

Maximum likelihood estimation of parameters in models with
discrete random effects requires removing them by
summation. For Markovian models, this can be done recursively,
making the likelihood tractable. A general expression of the likelihood
can be written:
\begin{multline}
  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
%  \prod_{t=T}^2 \left\{
  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
  \right\} \right\}
  \label{eq:like}
\end{multline}
where $\infty$ is replaced by a finite upper bound $K$, which
%%% Can we rename the upper bound?  We're also using K
%%% for equilibrium abundance.
should be high enough such that the MLEs are not affected by increasing
it further. This likelihood can be maximized numerically using the
\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
are given in the Appendix.

@@

We thank J. Andrew Royle and T. Scott Sillett for helpful
comments and suggestions. We gratefully acknowledge National Park 
Service, Smithsonian Institution, and [more] for funding support.
